\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-05-21 paper/subdocuments/abstract.Rnw
Test
some notes and comments
update of abtracts and introduction

\begin{abstract}

There is empirical evidence that economic inequality increased in the majority of western countries over the last decades (OECD, 2011; Gornick and Jäntti, 2013). In Switzerland, however, the development is unclear, as there is only little systematic evidence that would allow for long-term comparison. Nevertheless income inequality has been a prominent theme in the public discussion in recent years, e.g. the most recent referendums "Abzockerinitiative'' and "1:12-Initiative''. We adress the gap by presenting a long and consistent time series of inequality measures for income and wealth for Switzerland (1943-2010) calculated from federal tax data. We describe the benefits and shortcomings of tax data compared to other data sources and present strategies to handle tax data specific methodological difficulties. In the end we integrate the case of Switzerland into the international picture of inequality development showing parallels and deviations.

@@

>>>>

\begin{abstract}
% There is empirical evidence that economic inequality increased in the majority of western countries over the last decades (OECD, 2011; Gornick and Jäntti, 2013). In Switzerland, however, the development is unclear, as there is evidence for trends in both directions. Part of the inconclusive picture is due to different methodological approaches. In this paper we discuss the role of taxdata concerning the assessment of inequality in income. The focus of the discussion lays herein to show the benefits and shortcomings of tax data compared to current “state of the art” measurement concepts of economic inequality. We present common and new strategies to handle tax data specific methodological difficulties and compare results out of aggregated federal tax statistics to results from the Household Budget Survey (HBS). We can show to which extend survey data underestimate inequality in income. Following the results out of the taxdata Switzerland experienced in slight rise in inequality in recent years, similar to other western countries, but only because of rise in upper percentiles of the income distribution.

% Salverda-Stuff und das neue Piketty Buch sollten irgendwo erwähnt werden

@@

% Wir machen nichts mehr zu Vermögen (obwohl der Trend auch dahin geht, dass neben dem Einkommen auch das Vermögen eine zentrale Untersuchunsgrösse des inviduellen Wohlstandss darstellt)

%

% Alte version des abstractes. Fokus stärker auf Schweiz und Vermögen
%There is empirical evidence that economic inequality increased in the majority of western countries over the last decades (OECD, 2011; Gornick and Jäntti, 2013). In Switzerland, however, the development is unclear, as there is only little systematic evidence about income inequality that would allow for long-term comparison. Studies on the distribution of wealth are even scarcer. Nevertheless, income inequality has been a prominent theme in the public Swiss discussion in recent years, e.g. the most recent referendums "Abzockerinitiative'' and "1:12-Initiative''. We address the mentioned gap by presenting a long and consistent time series of inequality measures for income and wealth (1943-2010) calculated from federal tax statistics. We describe the benefits and shortcomings of tax data compared to other data sources and present strategies to handle tax data specific methodological difficulties. In the end we integrate the case of Switzerland into the international picture of inequality development showing parallels and deviations.

### 2014-05-21 paper/subdocuments/introduction.Rnw
update of abtracts and introduction

% Im European Sociologica Review scheint es unüblich zu sein, ohne theoretische Sektion zu arbeiten. Ich hoffe nicht, dass dies ein k.o. Kriterium ist. wir halten dafür die Einführungs und die Methoden Sektion etwas allgemeiner

@@

Economic Resources can be seen as central indicator for life chances in general and a multitude of outcomes like physical and mental health, life expectancy and crime in particular (Wilkinson and Pickett 2009). While the study of social inequality can be considered as one of the core subjects of sociology in more recent years the concern about the widening gap was addressed by global leaders (WEF 2013) and scholars alike. Empirical evidence acknowledge the supposed trend that economic inequality increased in the majority of western countries over the last decades (OECD, 2008, 2011; Gornick and Jäntti, 2013). The pattern of evolution suggests a polarization, which can be referred to as the "hollowing of the the middle class" (Alderson and Doran 2013). Households are moving towards the top and the bottom of the distribution relative to the past, which is especially problematic as the middle class can be seen as the core of western democracies or as it is stated by Stiglitz (2012,117): "our democracy is being put at peril."
\\

@@

% 3.Abschnitt: What do we know about Switzerland
%

What is known about Switzerland so far? Among the European Nations Switzerland can be situated in the middle-field when looking at inequality of disposable income (EU-SILC,Eurostat (2013)). Nevertheless in recent years Switzerland experienced  an ongoing political debate about the distribution of income in general and the widening of the gap between low and top earners. The debate went along with referendums trying to regulate the market outcomes. Whereas the "Rip-Off-Initiative" (Abzockerinitiative) found a surprisingly high majority of 68\%, the 1:12-Initiative, which aimed at the whole spectrum of the income distribution, was rejected. The voting about the minimum wage-initiative will be held on 9th February 2014. All these referendums questioned the current distribution of income and were accompanied by a broader discussion about the development of income inequality held in the media. Several official and semi-official publications addressed the question of rising inequality in Switzerland. \\

% Lässt sich allenfalls streichen/kürzen
% Ist es die Bezeichnung offical publications sinnvoll? Oder besser govermental publications
% rf: oder "Federal publications"?

% 4.Abschnitt: What do we know about Switzerland
% Offical Data Collections on development of inequality

The most recent official figures published by the Swiss Federal Statistical Office (2013) show stability concerning the income distribution covering the time period of 1998 until 2011. The authors found a slight increase in the primary income which endured until the recent crisis. This temporary increase was mainly compensated by governmental redistribution. A longer period is covered in the LIS-Data-set (1982-2004).\footnote{Data-provider for the LIS Data is the Swiss Federal Statistical Office. The data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000,2002,2004). The OECD-Database includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). This change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). All in all the LIS dataset contains the longest time series on inequality for Switzerland.} Analyzing these data Gornick and Jäntti (2013) found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by Grabka and Kuhn (2012) analyzing the Swiss Household Panel (2000-2009). \\

% 5.Abschnitt: What do we know about Switzerland
% Studies with tax data

Whereas the aforementioned publications focused on disposable household income from survey data, the revival of tax-data-inequality studies lead to fruitful insights for Switzerland as well. Dell et al. (2007) used tax data from the Federal Tax Administration to assess the development of concentration of the highest incomes and wealth (top-shares). In contrast to most other examined countries, Switzerland did not experience a reduction in income and wealth concentration from the pre-First World War period to the decades following the second World War (up to 1996). Using the same approach Foellmi and Martinez (2013) expand the Dell et al. timeline to 2008 finding that the share of top income has risen, the top 0.01\% share even doubled in the last observed 20 years. A result which opposes the outcome of official data published by the Swiss Federal Statistical Office.\footnote{There are other studies on Switzerland covering different periods but not the recent years. Flückiger et al. (2007) and also Jeitziner and Peters (2007,2009) report constant inequality from 1960-1996 respectively from 1995 to 2003. Covering a similar time period Bauer and Spycher (1994) and Bolzani and Abul Naga (2002) found decreasing inequality. On the other hand Buchmann and Sacchi (1995) and Ernst et al. (2009) found an increase in the 1980s).} \\

% 6. Abschnitt: Zwischenbilanz und Herausforderungen

@@

% Allenfalls hier noch nicht so ausführlich mit Vor- und Nachteilen von Steuerdaten kommen. Vielleicht besser darauf hinweisen, welche Dimensionen betroffen sind und das dies in der Methods Section besprochen wird.
%rf: finde den abschnitt super

While there are several studies about income inequality in Switzerland, the publications concerning the distribution of wealth are scarcer, albeit the distribution of wealth seems to be a relevant dimension shaping the economic well-being of individuals as well. The cross-national Data center in Luxembourg is expanding their efforts, constructing the first cross-national wealth database. But, up to the day information for Switzerland is not available. 
% ab hier wird es irgnediwe unklar
Shorrock et al. (2013) unify several datasources (including tax data) to assess the pattern of global wealth. Their databook includes figures for Switzerland. In contrast to inequality of income concerning inequality of wealth Switzerland takes a leading position \\

% the credit suisse wealth report already uses wealth tax records, but they use it only for the period from 2000-2011. They did exactly the same for Denmark and Norway, which could be interesting to compare.
% Erbschaftssteuer erwähnen und bei den Global Wealth Reports der Credit Suisse und den Studien im Forschungsantrag weitermachen.

Up to the day, Switzerland can be situated according to the actual level of income inequality in western societies as there is a huge effort to collect data which can be harmonized to comparable measures (see Luxembourg income study). However, it is unclear how the bias through non-response affects the overall measure of inequality. Likewise a long and consistent time-series allowing to identify and explore development patterns on every point of the distribution (not only the top-shares) and a time series for inequality of wealth is missing. Building on recent developments in the field of inequality research, we assess the suitability of the publicly accessible tax data to report inequality and its changes over time. First of all this includes a discussion of the accessible measures in context of a reflection about the state of the art conceptualization of economic resources as an indicator for economic well-being. Second, we summarize and apply tax data specific technics to construct suitable measures of inequality. We expand the given set by an in depth discussion of a newly applied step to handle the incomplete coverage in tax data statistics. Third, we integrate the development of Swiss inequality into the international picture.

% Should we discus the non-scientic literature? -> allgemeine Frage und im konkreten betrifft dies die Publikationen von Economie Swiss und des Gewerkschaftsbunds.
% rf: gute frage, spontan würde ich sagen eher nein, müssten wir mal diskutieren

>>>>

@@

Economic Resources can be seen as central indicator for life chances in general and a multitude of outcomes like physical and mental health, life expectancy and crime in particular (Wilkinson and Pickett 2009). While the study of social inequality can be considered as one of the core subjects of sociology in more recent years the concern about the widening gap was addressed by global leaders (WEF 2013) and scholars alike. Empirical evidence acknowledge the supposed trend that economic inequality increased in the majority of western countries over the last decades (OECD, 2008, 2011; Gornick and Jäntti, 2013;Salverda et al. 2014). Although the rise was not uniform, a common pattern seems to be identifiable, which can be referred to as the "hollowing of the the middle class" (Alderson and Doran 2013). Households are moving towards the top and the bottom of the distribution relative to the past, which is especially problematic as the middle class can be seen as the core of western democracies or as it is stated by Stiglitz (2012,117): "our democracy is being put at peril."
\\

% Allenfalls könnte man auch (Nolan und Whealan, 2014) erwähnen. "The social Impact of income inequality: Poverty, Deprivation, and Social Cohesion" Da werden die Thesen rund um die Folgen von Ungleichheit gut zusammengefasst. Wobei: Jetzt scheint es mir jetzt ziemlich knackig und flüssig

@@

% 3.Abschnitt: What do we know about Switzerland I
% Offical Data Collections on development of inequality

As we focus our paper on the case of Switzerland, it is important to embed our work in the context of given publications concering inequality in income. What is known about Switzerland so far? Looking for offical data concering income inequality in Switzerland, three main sources has to be mentioned, which can be considered as de facto offical data sources: EU-SILC, HBS and LIS-data. Figure XY shows the Results steming from this three sources while looking at Gini of equivalised disposable income.  Up to the day, EU-SILC or Statistics on Income and Living Conditions is the main source used for policy monitoring at EU-level. The main focus of EU-SILC is to collect data on a common "framework" to ensure comparability among EU-countries and countries living around or within the EU. As a Non-EU member Switzerland implemented the instrument not from the beginning (2004) but as from 2007. Therefore this times-series doesn't cover timepoints before 2007. As graph XY shows, following the results from SILC income inequality decreased from 2007 to 2013.\footnote{Data shown in the graph was downloaded from the Eurostat Metadata-portal http://epp.eurostat.ec.europa.eu/cache/ITY_SDDS/EN/ilc_esms.htm last accessed 21.Mai 2014.} The second important source concering the distribution of income is the Household Budget Survey (HBS). The main focus of this survey lays in providing detailed data on houshold budegts allowing to apply different income definitions. Since 2000 the survey has been conducted on a continuous basis, which allows to look at a consistend time series from 2000 to 2011. As it can be seen from graph XY the trend is rather stable.\footnote{figures shown in the graph were calculated out of the original datasets, which were kindly provided by Swiss Federal Statistical Office.} Both time-series (SILC and HBS) cover a relatively short time period. A longer period is covered in the LIS-Data-set (1982-2004). Data-provider for the LIS Data is the Swiss Federal Statistical Office. In contrast the the aformentioned surveys the LIS-data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000,2002,2004).  All in all the LIS dataset contains the longest time series on inequality for Switzerland. Analyzing this data Gornick and Jäntti (2013) found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by Grabka and Kuhn (2012) analyzing the Swiss Household Panel (2000-2009).\footnote{A further offical database for income distribution is the OECD-Database. It includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). But this change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). A further important database on inequality is the GINI database which have been derived from the GINI Country Reports (Nolan et al.,2014). But this dataset doesn't cover Switzerland.  \\

<<Swiss inequality from other studies>>=
library(ggplot2)
library(reshape)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
swissGini_long<-melt(swissGini,id="Year")
ggplot(data=swissGini_long,
  aes(y=value,x=Year),shape=variable)+
  ylab("Gini of equivalised disposable income") +
  scale_y_continuous(limits=c(0.18,0.4)) +
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=variable),size=3)+
  geom_line(data=swissGini_long[!is.na(swissGini_long$value),],aes(linetype=variable))+
  theme_bw()
@

% Achtung: die HABE Zeitreihe muss upgedatet werden. Ist jetzt von Auge aus dem Bericht abgeschrieben.

% 4.Abschnitt: What do we know about Switzerland II
% Studies with tax data

Whereas the aforementioned publications focused on equivalised disposable income from survey data, the revival of tax-data-inequality studies lead to fruitful insights for Switzerland as well. Dell et al. (2007) used tax data from the Federal Tax Administration to assess the development of concentration of the highest incomes and wealth (top-shares). In contrast to most other examined countries, Switzerland did not experience a reduction in income and wealth concentration from the pre-First World War period to the decades following the second World War (up to 1996). Using the same approach Foellmi and Martinez (2013) expand the Dell et al. timeline to 2008 finding that the share of top income has risen, the top 0.01\% share even doubled in the last observed 20 years. A result which opposes the outcome of official data published by the Swiss Federal Statistical Office. Furthermore Schaltegger and Gorgas (2011) found a very heterogeneous evolution on the sub-federal level due to decentralized taxing power. \footnote{There are other studies on Switzerland covering different periods but not the recent years. Flückiger et al. (2007) and also Jeitziner and Peters (2007,2009) report constant inequality from 1960-1996 respectively from 1995 to 2003. Covering a similar time period Bauer and Spycher (1994) and Bolzani and Abul Naga (2002) found decreasing inequality. On the other hand Buchmann and Sacchi (1995) and Ernst et al. (2009) found an increase in the 1980s).} \\

% 6. Abschnitt: Zwischenbilanz und Herausforderungen. Datenlage.

@@

% Zusammenfassung Stand der Forschung und festlegen unserer Ausrichtung

Up to the day, Switzerland can be situated according to the actual level of income inequality in western societies as there is a huge effort to collect data which can be harmonized to comparable measures (see Luxembourg income study, EU-SILC). However, it is unclear how the bias through non-response affects the overall measure of inequality. Likewise a long and consistent time-series allowing to identify and explore development patterns on every point of the distribution (not only the top-shares) is missing. Building on recent developments in the field of inequality research, we assess the suitability of the publicly accessible tax data to report inequality and its changes over time. First of all this includes a discussion of the accessible measures in context of a reflection about the state of the art conceptualization of economic resources as an indicator for economic well-being. Second, we summarize and apply tax data specific technics to construct suitable measures of inequality. We expand the given set by an in depth discussion of a newly applied step to handle the incomplete coverage in tax data statistics. Third, we compare the results with the HBS time series to get an idea of the survey bias throgh non-response.

% Was wollen wir wissen?
% Wir müssen uns äussern zu: 
% Definition des Einkommens
% Analyseinheiten
% Berechnungsschritten

%%%%
% 0) In der Methods-Sektion beschreiben was wir machen? So was wie eine Forschungsstrategie/Vorgehensweise
% Wie ausführlich?
% Gini-Koeffizient, Relative Distribution-Methods. Berechnung?

% Bei den Ergebnissen müsste es Grafiken haben zu:
% 1) Unterschiedlichen Analyseneinheiten (Mit und ohne Nuller, Normal- und Sonderfälle)
% 2) Unterschiedliche Einkommensgrössen (Reineinkommen, steuerbares Einkommen, pseudoverfügbares Einkommen (Abzüge von direkten Bundessteuern))
% 3) alternative measure of income inequality: relative distribution and polaritätsindex
% 4) Vergleich mit HABE

%%%%%%%%%%%
% Alte Textbausteine

% Textbaustein, zur politischen Dimension von Einkommensungleichheit (alt)
%Nevertheless in recent years Switzerland experienced  an ongoing political debate about the distribution of income in general and the widening of the gap between low and top earners. The debate went along with referendums trying to regulate the market outcomes. Whereas the "Rip-Off-Initiative" (Abzockerinitiative) found a surprisingly high majority of 68\%, the 1:12-Initiative, which aimed at the whole spectrum of the income distribution, was rejected. The voting about the minimum wage-initiative will be held on 9th February 2014. All these referendums questioned the current distribution of income and were accompanied by a broader discussion about the development of income inequality held in the media. Several official and semi-official publications addressed the question of rising inequality in Switzerland. \\

% Textbaustein (unfertig) zu Vermögensungleichheit (alt)
% While there are several studies about income inequality in Switzerland, the publications concerning the distribution of wealth are scarcer, albeit the distribution of wealth seems to be a relevant dimension shaping the economic well-being of individuals as well. The cross-national Data center in Luxembourg is expanding their efforts, constructing the first cross-national wealth database. But, up to the day information for Switzerland is not available. Shorrock et al. (2013) unify several datasources (including tax data) to assess the pattern of global wealth. Their databook includes figures for Switzerland. In contrast to inequality of income concerning inequality of wealth Switzerland takes a leading position \\

% Auch erste Dell Studie erwähnen -> es hat Zahlen zur Schweiz. Auch Probleme erwähnen (Fehlende Informationen zu Altersvorsorgevermögen -> hat aber niemand eine Lösung bisher.)
